ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-10 notebooks/tables.Rmd
just results, limitations and conclusion left

@@

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="real",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,std:=NULL]
tab[,min_val:=NULL]
tab[,two_std_wachter:=NULL]
tab[,one_std_wachter:=NULL]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for real-world datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-real} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_real.tex")
```

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="synthetic",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,std:=NULL]
tab[,min_val:=NULL]
tab[,two_std_wachter:=NULL]
tab[,one_std_wachter:=NULL]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Non-conformity ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for synthetic datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-synth} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "Circles" = 2, "Linearly Separable" = 2, "Moons" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_synth.tex")
```

>>>>

```{r}
tab <- dt[
    ,
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable, source)
]

tab$top_val = F
tab$one_std_wachter = F
tab$two_std_wachter = F

# Measures to be minimized:
min_measures <- c(
  "distance",
  "distance_from_energy", 
  "distance_from_targets",
  "set_size_penalty"
)
tab[variable %in% min_measures,top_val:=val==min(val),.(model, dataname, variable)]
tab[variable %in% min_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab[variable %in% min_measures,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[variable %in% min_measures,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]

# Measures to be maximized:
max_measures <- c(
  "validity",
  "redundancy"
)
tab[variable %in% max_measures,top_val:=val==max(val),.(model, dataname, variable)]
tab[variable %in% max_measures,top_val:=ifelse(rep(all(top_val),length(top_val)),F,top_val),.(model, dataname, variable)]
tab[variable %in% max_measures,two_std_wachter:=val-2*std>val[generator=="Wachter"],.(model, dataname, variable)]
tab[variable %in% max_measures,one_std_wachter:=val-1*std>val[generator=="Wachter"],.(model, dataname, variable)]

# Add conditional formatting:
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$top_val)

# Remove redundant columns:
tab[,val:=NULL]
tab[,std:=NULL]
tab[,top_val:=NULL]
tab[,two_std_wachter:=NULL]
tab[,one_std_wachter:=NULL]
```

## Full table

```{r}
tab_full <- dcast(tab, dataname + model + generator ~ variable)
col_names <- c(
  "Model",
  "Data",
  "Generator",
  "Cost ↓", 
  "Unfaithfulness ↓", 
  "Implausibility ↓", 
  "Redundancy ↑",
  "Uncertainty ↓",
  "Validity ↑"
)
kbl(
  tab_full, caption = "All results for all datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-full} \\newline", 
  align = "c", col.names=col_names, booktabs = F, escape=F, 
  format="latex"
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:3, latex_hline = "custom", valign = "middle", custom_latex_hline = 1:2) %>%
  save_kable("paper/contents/table_all.tex")
```

@@

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="real",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,std:=NULL]
tab[,min_val:=NULL]
tab[,two_std_wachter:=NULL]
tab[,one_std_wachter:=NULL]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Unfaithfulness ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for real-world datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-real} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "California Housing" = 2, "GMSC" = 2, "MNIST" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_real.tex")
```

```{r}
measures <- c("distance_from_energy", "distance_from_targets")
tab <- dt[
    variable %in% measures & source=="synthetic",
    .(
      value=sprintf("%1.2f (%1.2f)", mean(value), sd(value)),
      val = mean(value),
      std = sd(value)
    ),
    .(dataname, generator, model, variable)
]
tab[,min_val:=val==min(val),.(model, dataname, variable)]
tab[,two_std_wachter:=val+2*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[,one_std_wachter:=val+1*std<val[generator=="Wachter"],.(model, dataname, variable)]
tab[one_std_wachter==T,value:=paste0(value,"*")]
tab[two_std_wachter==T,value:=paste0(value,"*")]
tab$value <- cell_spec(tab$value, "latex", bold=tab$min_val)
tab[,val:=NULL]
tab[,std:=NULL]
tab[,min_val:=NULL]
tab[,two_std_wachter:=NULL]
tab[,one_std_wachter:=NULL]
n_data <- length(unique(tab$dataname))
tab <- dcast(tab, model + generator ~ dataname + variable)
col_names <- c(
  "Model",
  "Generator",
  rep(c("Unfaithfulness ↓","Implausibility ↓"),n_data)
)
kbl(
  tab, caption = "Results for synthetic datasets. Standard deviations across samples are shown in parentheses. Best outcomes are highlighted in bold. Asterisks indicate that the given value is more than one (*) or two (**) standard deviations away from the baseline (Wachter). \\label{tab:results-synth} \\newline", 
  align = "c", col.names=col_names, booktabs = T, escape=F, 
  format="latex", linesep = c("", "", "", "\\addlinespace") 
) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  kable_paper(full_width = F) %>%
  add_header_above(c(" " = 2, "Circles" = 2, "Linearly Separable" = 2, "Moons" = 2)) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, latex_hline = "major", valign = "middle") %>%
  save_kable("paper/contents/table_synth.tex")
```

### 2023-05-10 paper/paper.tex
just results, limitations and conclusion left

In the following, we first briefly describe our evaluation framework and data, before presenting and discussing our results.

@@

As noted by~\citet{guidotti2022counterfactual}, these distance-based measures are simplistic and more complex alternative measures may ultimately be more appropriate for the task. For example, we considered using statistical divergence measures instead. This would involve generating not one but many counterfactuals and comparing the generated empirical distribution to the target distributions in Definitions~\ref{def:plausible} and~\ref{def:faithful}. While this approach is potentially more rigorous, generating enough counterfactuals is not always practical.

\subsection{Data}

@@

\subsection{Results}\label{app:results}

>>>>

In the following, we first briefly describe our evaluation framework and experimental setup, before presenting and discussing our results.

@@

\subsection{Experimental Setup}

To assess and benchmark the performance of ECCCo against the state of the art, we generate multiple counterfactuals for different black-box models and datasets. In particular, we compare ECCCo to the following counterfactual generators that were introduced above: firstly; \textbf{Schut}~\citep{schut2021generating}, which minimizes predictive uncertainty; secondly, \textbf{REVISE}~\citep{joshi2019realistic}, which uses a VAE as its surrogate model; and, finally, \textbf{Wachter}~\citep{wachter2017counterfactual}, which serves as our baseline.

We use both synthetic and real-world datasets from different domains, all of which are publically available. The synthetic datasets include: a dataset containing two \textbf{Linearly Separable} Gaussian clusters ($n=1000$), as well as the common \textbf{Circles} ($n=1000$) and \textbf{Moons} ($n=2500$) data. As for real-world data, we use two well-known datasets from the vision domain: firstly, following~\citet{schut2021generating} we use \textbf{MNIST}~\citep{lecun1998mnist} and, secondly, we add \textbf{Fashion-MNIST}~\citep{xiao2017fashion}. Both datasets involve 60000 images of 28x28 pixels each in 10 different classes. Finally, we also study the Give Me Some Credit (\textbf{GMSC}) dataset, a tabular dataset from the social sciences domain that has been used extensively in the literature on Algorithmic Recourse~\citep{pawelczyk2021carla}. It consists of 11 numeric features that can be used to predict the binary outcome variable indicating whether or not retail borrowers experience financial distress~\citep{kaggle2011give}.

As for the example in Section~\ref{meth}, we use simple neural networks (\textbf{MLP}), ensembles of neural networks (\textbf{MLP Ensemble}), Joint Energy Models (\textbf{JEM}) and ensembles of JEMs (\textbf{JEM Ensemble}) to model our real-world datasets. For the synthetic datasets, we found that the use of ensembles was not necessary. Full details concerning our hyperparameter choices, training procedures and model performance can be found in Appendix~\ref{app:setup}.

@@

\subsection{Experimental Setup}\label{app:setup}
\subsection{Results}\label{app:results}

\import{contents/}{table_all.tex}
